\Huge State-Space Models and Particle Filters

### 2021-07-30 backup1.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

  pdftitle={Final draft},

@@

\title{Final draft}
\author{}
\date{\vspace{-2.5em}}

@@

\chapter{Markov Processes and State-Space Models}
\section{Notation}

@@

Let \(P_1, P_2, \dots, P_T\) be a finite sequence of probability kernels
from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)

@@

It is possible to show that for any \(t\in T,\) \[
\mathbb{P}_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\mathbb{P}_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\mathbb{P}_t(x_{t-1},dx_t).
\] The first equality implies conditional independence. The second
equality implies that it is possible to identify conditional
distributions with transition kernels. Assume \(\mathbb{P}_t,\) with
\(t\leq T,\) is a sequence of probability measures. Then, the following
proposition holds.

@@

We now define \textit{State-Space models} as Markov processes that are
partially observed. Let \(X_t\in\mathcal X\) and \(Y_t\in\mathcal Y\)
and let \(\{(X_t,Y_t)\}\) be a stochastic process. To help intuition,
suppose that \(\{Y_t\}\) is observable, i.e., coincides with a sequence
of observations, while \(\{X_t\}\) is latent. Consider the initial
distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).

@@

Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
\(\mathcal Y=\mathbb R^{d_y}.\) Let
\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
and assume the initial distribution
\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
if \begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
normally distributed.

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-5-1}

@@

\hrule
\hrule
\texttt{DLM(data,sig2,tau2,m0,C0)}
\hrule

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-7-1}

@@

Let \(\varphi: \mathcal X\rightarrow \mathbb R\) be a measurable,
integrable and bounded real-valued function, where
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) is a probability
space.\footnote{$\mathcal B(\mathcal X)$ is the Borel sigma-algebra of $\mathcal X$.}
Suppose that the probability measure \(\mathbb Q\) admits a density
\(q(x)\) with respect to some measure \(\nu(dx)\) and it is feasible to
sample from that density.\\
To estimate the expected value of \(\varphi\) with respect to the
density \(q\), the basic Monte Carlo (MC) method would require to sample
\(N\) times from the target density \(q\) and to compute the average of
the function \(\varphi\) values. \(\hat \varphi_{MC}\) is the standard
MC estimator: \begin{equation*}
   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \mathbb E_q(\varphi)~~~X^n\thicksim q(\cdot)
\end{equation*} However, in many circumstances either it is impossible
to sample from the target density or there exist another density - call
it \(m(\cdot)\) - different from \(q\) that produces
\texttt{more\ efficient"\ estimates.\ \ In\ all\ the\ cases\ for\ which\ sampling\ takes\ place\ from\ a\ distribution\ different\ from\ the}correct"
(i.e., the true) one, importance sampling and resampling methods can be
applied (as augmented versions of the standard MC algorithm).

@@

Usually, PF requires a resampling step at each stage of the algorithm.
Recall that, in general, resampling determines a trade-off between the
increase in the variance at the resampling stage and a decrease in the
variance when future states are sampled from the transition kernel.\\
\texttt{Dead"\ particles\ (i.e.,\ those\ with\ small\ weights)\ are\ \ likely\ to\ be\ removed\ when\ resampling\ takes\ place\ and\ past\ weights\ get}reset"
(usually, they are set to be equal to \(1/N\)). Otherwise, the
\textit{curse of dimensionality} problem will emerge and the variance of
the weights will diverge over time as the variance of past weights
increases as new updates occur period-by-period. Figure ***** shows an
example of a PF where SIS is used: note that the \(ESS\) drops
dramatically after few periods, meaning that the variance of the
importance weights diverges. This motivates the use of a SIR structure
for the PF algorithm.~\\
Lastly, it is now common practice to use a mixed approach with respect
to resampling (recall that resampling has an additional computational
cost). Generally, resampling is done \textit{adaptively}, when some
condition is met. A standard criterion that is used (and that we adopt
in our implementations of the PF) is to resample when the \(ESS\) drops
below some threshold \(ESS_{min}\in [1,N]\): \begin{equation}
    ESS(W_{t-1}^{1:N})<ESS_{min}
\end{equation}

\section{Implementation}

@@

\hrule
\hrule
\texttt{SISfun(data,N,m0,C0,tau,sigma)}
\hrule

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-10-1}

@@

\hrule
\hrule
\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}
\hrule

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-13-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-15}Root Mean Square Errors}\\
\toprule
N & Threshold & KF & BPF\\
\midrule
100 & 0.5 & 0.879 & 0.888\\
1000 & 0.5 & 0.879 & 0.886\\
10000 & 0.5 & 0.879 & 0.878\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-16-1}

@@

\hrule
\hrule
\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}
\hrule

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-19-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-21}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & GPF\\
\midrule
1000 & 0.50 & 0.916 & 0.880\\
1000 & 0.25 & 0.895 & 0.862\\
1000 & 0.10 & 0.869 & 0.881\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-22-1}

@@

For illustration purposes, we are going implement an auxiliary particle
filter with auxiliary function
\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).\\

@@

\hfill\break
\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-25-1}

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-26-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-28}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & APF\\
\midrule
1000 & 0.50 & 0.885 & 0.875\\
1000 & 0.25 & 0.893 & 0.892\\
1000 & 0.10 & 0.855 & 0.870\\
\bottomrule
\end{longtable}

@@

@@

{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-31-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-31}
\end{figure}

@@

Let us consider a more general state-space model, where the state vector
includes a time-constant parameter vector \(\psi \in \Psi\), where
\(\Psi\) is the parameter space. Since we can interpret the parameter as
a state with the law of motion \(\psi_t=\psi_{t-1}=\psi\), then it is
possible to adopt a PF algorithm to estimate \(\psi\).\\
However, being \(\psi\) constant over time, it is meaningless to sample
its values sequentially: the first extraction of \(\psi\) at the initial
stage 0 (i.e., \(\psi_0^n\)) generates a constant path of particles for
the parameter (\(\psi_t^n=\psi_0^n\)).
\footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\\
Therefore, Liu and West (2001) proposed a modified particle filter,
called \textit{Liu and West filter} (LWF) that allows to resample the
parameter over time from a continuous distribution. In this way, at
every time, the support of \(\psi\) is not limited to the initially
sampled \(N\) values.\\
Here, we describe the LWF that makes use of the normal distribution to
sample sequentially the parameter. Moreover, consider the simpler
version of the bootstrap
filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\\
Let the transition kernel and the likelihood depend on the parameter
\(\psi\) (i.e., \(P_t(x_{t-1},dx_t;\psi)\) and \(f_t(y_t|x_t;\psi)\))
and let \$\pi(\cdot) \$ be the prior distribution for the parameter
\(\psi\).\\
The LWF can be described by the following algorithm.

@@

@@

{\centering \includegraphics{new-draft_files/figure-latex/cll_plot-1}

>>>>

  pdftitle={State-Space Models and Particle Filters},
  pdfauthor={Edoardo Marcelli; Roberto Colarieti; Alexei Verkhovtsev; Michele Bolla; Gabriele Romano; Benedetta Bruni},

@@

\title{\Huge State-Space Models and Particle Filters}
\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei
Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta
Bruni}
\date{21 luglio, 2021}

@@

\chapter{Markov Processes and State-Space Models}

\section{Notation}

@@

Let \(P_1, P_2, \dots, P_T\) be a finite sequence of probability kernels
from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)\\

@@

It is possible to show that for any \(t\in T,\) \[
\mathbb{P}_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\mathbb{P}_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\mathbb{P}_t(x_{t-1},dx_t).
\] The first equality implies conditional independence. The second
equality implies that it is possible to identify conditional
distributions with transition kernels. Assume \(\mathbb{P}_t,\) with
\(t\leq T,\) is a sequence of probability measures. Then, the following
proposition holds.\\

@@

We now define \textit{State-Space models} as Markov processes that are
partially observed. Let \(X_t\in\mathcal X\) and \(Y_t\in\mathcal Y\)
and let \(\{(X_t,Y_t)\}\) be a stochastic process. To help intuition,
suppose that \(\{Y_t\}\) is observable, i.e., coincides with a sequence
of observations, while \(\{X_t\}\) is latent. Consider the initial
distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).\\

@@

Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
\(\mathcal Y=\mathbb R^{d_y}.\) Let
\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
and assume the initial distribution
\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
if \begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
normally distributed.\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-5-1}

@@

\hfill\break
\texttt{DLM(data,sig2,tau2,m0,C0)}\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-7-1}

@@

Let \(\varphi: \mathcal X\rightarrow \mathbb R\) be a measurable,
integrable and bounded real-valued function, where
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) is a probability
space.\footnote{$\mathcal B(\mathcal X)$ is the Borel sigma-algebra of $\mathcal X$.}
Suppose that the probability measure \(\mathbb Q\) admits a density
\(q(x)\) with respect to some measure \(\nu(dx)\) and it is feasible to
sample from that density.\\
To estimate the expected value of \(\varphi\) with respect to the
density \(q\), the basic Monte Carlo (MC) method would require to sample
\(N\) times from the target density \(q\) and to compute the average of
the function \(\varphi\) values. \(\hat \varphi_{MC}\) is the standard
MC estimator: \begin{equation*}
   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \mathbb E_q(\varphi)~~~X^n\thicksim q(\cdot)
\end{equation*} However, in many circumstances either it is impossible
to sample from the target density or there exist another density - call
it \(m(\cdot)\) - different from \(q\) that produces ``more efficient''
estimates. In all the cases for which sampling takes place from a
distribution different from the ``correct'' (i.e., the true) one,
importance sampling and resampling methods can be applied (as augmented
versions of the standard MC algorithm).

@@

Usually, PF requires a resampling step at each stage of the algorithm.
Recall that, in general, resampling determines a trade-off between the
increase in the variance at the resampling stage and a decrease in the
variance when future states are sampled from the transition kernel.\\
``Dead'' particles (i.e., those with small weights) are likely to be
removed when resampling takes place and past weights get ``reset"
(usually, they are set to be equal to \(1/N\)). Otherwise, the
\textit{curse of dimensionality} problem will emerge and the variance of
the weights will diverge over time as the variance of past weights
increases as new updates occur period-by-period. Figure ***** shows an
example of a PF where SIS is used: note that the \(ESS\) drops
dramatically after few periods, meaning that the variance of the
importance weights diverges. This motivates the use of a SIR structure
for the PF algorithm.~\\
Lastly, it is now common practice to use a mixed approach with respect
to resampling (recall that resampling has an additional computational
cost). Generally, resampling is done \textit{adaptively}, when some
condition is met. A standard criterion that is used (and that we adopt
in our implementations of the PF) is to resample when the \(ESS\) drops
below some threshold \(ESS_{min}\in [1,N]\): \begin{equation}
    ESS(W_{t-1}^{1:N})<ESS_{min}
\end{equation}

\subsection{Implementation}

@@

\hfill\break
\texttt{SISfun(data,N,m0,C0,tau,sigma)}\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-10-1}

@@

\hfill\break
\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-13-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-15}Root Mean Square Errors}\\
\toprule
N & Threshold & KF & BPF\\
\midrule
100 & 0.5 & 0.879 & 0.916\\
1000 & 0.5 & 0.879 & 0.882\\
10000 & 0.5 & 0.879 & 0.885\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-16-1}

@@

\hfill\break
\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-19-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-21}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & GPF\\
\midrule
1000 & 0.50 & 0.882 & 0.874\\
1000 & 0.25 & 0.889 & 0.874\\
1000 & 0.10 & 0.885 & 0.879\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-22-1}

@@

\subsection{Implementation}

For illustration purposes, we are going implement an auxiliary particle
filter for the linear Gaussian model with auxiliary function
\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).

@@

\hfill\break
\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}\\

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-25-1}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-26-1}

@@

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-28}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & APF\\
\midrule
1000 & 0.50 & 0.882 & 0.899\\
1000 & 0.25 & 0.889 & 0.883\\
1000 & 0.10 & 0.885 & 0.877\\
\bottomrule
\end{longtable}

We also implement the auxiliary particle filter with optimal transition
kernel. Therfore the algorithm is similiar to the one discussed above
with the difference that:\\
{[}{[}{[}{[}Inserire l'algoritmo{]}{]}{]}{]}

@@

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
\DocumentationTok{\#\# Per roberto:::::: cambia molto se usiamo l\textquotesingle{}uno o l\textquotesingle{}altro!!!!!! perchè??}
\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}

\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-32-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
\end{figure}

@@

Let us consider a more general state-space model, where the state vector
includes a time-constant parameter vector \(\psi \in \Psi\), where
\(\Psi\) is the parameter space. Since we can interpret the parameter as
a state with the law of motion \(\psi_t=\psi_{t-1}=\psi\), then it is
possible to adopt a PF algorithm to estimate \(\psi\).\\
However, being \(\psi\) constant over time, it is meaningless to sample
its values sequentially: the first extraction of \(\psi\) at the initial
stage 0 (i.e., \(\psi_0^n\)) generates a constant path of particles for
the parameter (\(\psi_t^n=\psi_0^n\)).
\footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\\
Therefore, Liu and West (2001) proposed a modified particle filter,
called \textit{Liu and West filter} (LWF) that allows to resample the
parameter over time from a continuous distribution. In this way, at
every time, the support of \(\psi\) is not limited to the initially
sampled \(N\) values.\\
Here, we describe the LWF that makes use of the normal distribution to
sample sequentially the parameter. Moreover, consider the simpler
version of the bootstrap
filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\\
Let the transition kernel and the likelihood depend on the parameter
\(\psi\) (i.e., \(P_t(x_{t-1},dx_t;\psi)\) and \(f_t(y_t|x_t;\psi)\))
and let \(\pi(\cdot)\) be the prior distribution for the parameter
\(\psi\).\\
The LWF can be described by the following algorithm.

@@

We provide an example of Liu and West filter fixing \(\delta=0.7\) and
drawing \(\alpha_{v}\), \(\beta_{v}\), \(\alpha_{w}\) and \(\beta_{w}\)
from a Uniform \(U(0,10)\). The results are shown in the following
figure.

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-35-1}

\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-35}
\end{figure}

Finally, we compare the performances of the Liu and West filter and the
other filtering strategies discussed so far. As we expect, in the linear
Gaussian case, Kalman filter outperforms the other strategies. On the
other hand, the Liu and West, with hyperparameters setted as described
above, provides the worst estimates in terms of RMSE. This is not
necessarily due to an overall weakness of the LWF with respect to the
other filters, but simply to a not accurate choice of the
hyperparameters. Moreover, while increasing the number of particles
improves the accuracy, notice that the threshold for ESS-based
resampling is not strictly related to better performances.

\begin{longtable}[t]{cccccc}
\caption{\label{tab:unnamed-chunk-37}RMSE}\\
\toprule
N & Threshold & KF & BPF & APF & LWF\\
\midrule
100 & 0.50 & 0.879 & 0.916 & 0.881 & 0.883\\
1000 & 0.50 & 0.879 & 0.882 & 0.899 & 0.977\\
10000 & 0.50 & 0.879 & 0.885 & 0.874 & 0.945\\
1000 & 0.50 & 0.879 & 1.172 & 1.677 & 1.140\\
1000 & 0.25 & 0.879 & 1.085 & 0.974 & 3.419\\
\addlinespace
1000 & 0.10 & 0.879 & 4.285 & 0.994 & 1.142\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/cll_plot-1}
